---
title: '[R] Use trelliscopejs to arrange plots'
author: "Zhenguo Zhang"
date: '2021-08-12'
slug: r-use-trelliscopejs-to-arrange-plots
categories:
- R
- Software
tags:
- blogdown
- R
- R Markdown
- plot
description: ''
featured_image: ''
images: []
comment: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Trelliscope is an R package which interfacing a Javascript library
[trelliscopejs](https://github.com/hafen/trelliscopejs-lib). This
library provides an visulization of data by splitting them into
subsets, also called 'Trellis Display' or "small multiples".

The R package provides some high-level functions such as
`facet_trelliscope()` which can directly replace the function
`ggplot2::facet_wrap()` when making plots. The advantage of the
former is that it can create multiple pages of plots when there
are many, but `facet_wrap()` will put them into one plot. Also
the trelliscope plot is an interactive htmlwidget, so users can
sort, filter, and arrange the plots.

Let's start with a demo.

## facet_trelliscope

Here we use the function `facet_trelliscope()` to wrap
around plots based on the data in `mtcars`.

```{r}
library(trelliscopejs)
library(ggplot2)
ggplot(data=mtcars, aes(x=wt,y=mpg)) + geom_point() +
  facet_trelliscope(~am + gear, nrow=1, ncol = 2,
                    name="Test trelliscope", desc="This is desc",
                    self_contained = F, path=".",
                    as_plotly = F)
```

As you can see, the usage is pretty straightforward. In the 
trelliscope panel, one can change the grid and sort the plots
based on precomputed metrics. Here we have only 4 plots, so
its power to organize many plots is not well demonstrated, but
the principle is the same.

Notes on the arguments of `facet_trelliscope()`:

* `~am + gear`: the formula to facet the panels. Check 
  `ggplot2::facet_wrap()` for more details.

* `nrow` and `ncol`: the number of rows and columns for
  the panel grid to initally display.
  
* `name` and `desc`: the main title and description for the
  trellis display.
  
* `self_contained`: set to TRUE when to include all the
  plots in the output html file; otherwise a folder 'appfiles'
  is produced to store the plots and is needed when browsing the
  output html file. However, if the plots are plotly format
  (when argument `as_plotly` is TRUE), this doesn't work and
  `self_contained` need be set to 'FALSE', as done here.

* `path`: this is the base directory for the trelliscope
  display to work; basically it relies on this to find 'appfiles'
  folder. This must be set to '.' when trelliscope display is
  embedded in Rmarkdown file, just like here.
  
* `as_plotly`: make plots as interactive plotly objects.

This function also have many other arguments, which can be
checked from its help page.


## References

1. Introduction to trelliscopejs: https://cran.r-project.org/web/packages/trelliscopejs/vignettes/trelliscopejs.html
