---
title: '[R] How to change legend key size in ggplot2'
author: Zhenguo Zhang
date: '2026-04-27'
slug: r-how-to-change-legend-key-size-in-ggplot2
categories:
  - R
  - Visualization
tags:
  - ggplot2
description: 'A quick tip on how to change the legend key size independently of text size in ggplot2.'
featured_image: ''
images: []
comment: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 5)
```

In `ggplot2`, the legend keys (the symbols next to the labels) usually inherit their size and appearance from the layers in the plot. For example, if you have a scatter plot with very small points, the legend keys will also be very small, making them hard to see.

To fix this without changing the size of the points in the actual plot, you can use the `override.aes` argument within `guide_legend()`.

### The Problem: Tiny Legend Keys

When plotting many data points, we often decrease the `size` to avoid overplotting. However, this makes the legend keys tiny:

```{r tiny-legend}
library(ggplot2)

p <- ggplot(iris, aes(Sepal.Length, Sepal.Width, color = Species)) +
  geom_point(size = 0.5) + 
  theme_minimal() +
  labs(title = "Tiny Legend Keys")

p
```

### The Solution: Using `override.aes`

By using `guides()`, we can "override" the aesthetics of the legend. This allows us to keep the small points in the plot while making the legend keys large and clear.

If your legend is for the **color** aesthetic:

```{r fixed-legend}
p_fixed <- p + 
  guides(color = guide_legend(override.aes = list(size = 5))) +
  labs(title = "Fixed Legend Keys (Size 5)")

p_fixed
```

If you were using **fill** (e.g., in a bar plot or boxplot), you would simply change `color` to `fill`:

```r
guides(fill = guide_legend(override.aes = list(size = 5)))
```

### Adjusting Legend Text and Title Size

Note that `override.aes` only affects the symbols inside the legend. To change the size of the labels or the legend title, you should use the `theme()` function:

```{r theme-adjust}
p_fixed + 
  theme(
    legend.text = element_text(size = 12),   # Legend labels
    legend.title = element_text(size = 14, face = "bold") # Legend title
  )
```

### Summary

*   Use `guides(color = guide_legend(override.aes = list(size = ...)))` to change the **symbol size**.
*   Use `theme(legend.text = element_text(size = ...))` to change the **text size**.
