---
title: '[R] data.table''s frank()'
author: Zhenguo Zhang
date: '2025-04-12'
slug: r-data-table-s-frank
categories:
  - R
  - Computing
tags:
  - R
description: ''
featured_image: ''
images: []
comment: yes
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(knitr)
library(data.table)
```

One can use `data.table::frank()` to rank the rows of a data.table or simply
a vector. Compared to the base R function `rank()`, `frank()` is faster. Today
I will show how to use this function.

First, let's generate a example data.table with 10 rows and 3 columns,
for simplicity, we will make first 2 columns are integer and the last one is
a character. Also, we will duplicate some values to show how tied values are
sorted:

```{r}
set.seed(123)
n <- 10
dt <- data.table(
  a = sample(1:10, n, replace = TRUE),
  b = sample(1:10, n, replace = TRUE),
  c = sample(letters[1:5], n, replace = TRUE)
)
kable(dt, caption = "Example data.table")
```

First, let's see how to use `frank()` to rank the whole data.table.

```{r}
dt[, rank := frank(.SD)]
kable(dt[order(rank)], caption = "Ranked data.table")
```

As you can see, the `frank()` function ranks the rows of the data.table
by first checking the first column, then the second column, and finally the third column.

One can also sort a data.table based on selected columns, for example,
let's use the 2nd and 3rd columns to rank the data.table. But for this,
one need to use its variant *frankv()*:

```{r}
dt[, rank := frankv(.SD, cols = c("b","c"))]
kable(dt[order(rank)], caption = "Ranked data.table by 2nd and 3rd columns")
```

Finally, we would like to talk about the `ties.method` argument. To make it simple,
we will simiply use the 2nd column to rank the table so you can see the effect of
the `ties.method` argument.

```{r}
newDT <- dt[, .(b)]
newDT[, rankAverage := frank(b, ties.method = "average")] # the default
newDT[, rankFirst := frank(b, ties.method = "first")]
newDT[, rankLast := frank(b, ties.method = "last")]
newDT[, rankRandom := frank(b, ties.method = "random")]
newDT[, rankMax := frank(b, ties.method = "max")]
newDT[, rankMin := frank(b, ties.method = "min")]
newDT[, rankDense := frank(b, ties.method = "dense")]
kable(newDT[order(b)], caption = "Ranked data.table by 2nd column")
```

As you can see, here are how the `ties.method` argument works:

- average: the average of the ranks of the tied values
- first: the order in which the values appear in the data
- last: the order in which the values appear in the data
- random: a random order for the ties
- max: the maximum rank of the tied values
- min: the minimum rank of the tied values
- dense: the values in a tie set get the same rank, and the rank
  value increases by 1 when moving to the next tie set. This is
  a unique feature of `frank()` and is not available in the base R.
  
When one wants to use the rank to choose top N rows, it is important
to know how the rank is computed; in this case, you may want to
avoid the `ties.method` values: `max`, `min`, and `dense`.

Happy programming :smile:

