---
title: '[R] Compute stats on grouped data using dplyr'
author: Zhenguo Zhang
date: '2023-05-27'
slug: r-compute-stats-on-grouped-data-using-dplyr
categories:
  - R
  - Computing
tags:
  - R
  - Programming
description: ''
featured_image: ''
images: []
comment: yes
---

```{r setup, warning=F, message=F}
library(knitr)
opts_chunk$set(echo=T, warning=T, fig.width=7, height=7)
library(dplyr)
```

[dplyr](https://dplyr.tidyverse.org/r) is a great package to process data
in a more smooth way: the output of one function can be inject to another
function using the operator `%>%` easily. Today, I would like to share my
experience on how to process grouped data and return multiple columns/rows
data.

## Data

We will generate some fake data for this test.

```{r}
g1<-rep(c("a","b","c"), each=10)
g2<-rep(c("A","B"), each=5, length.out=30)
x<-rnorm(30,0,10)
dat<-data.frame(g1,g2,x)
kable(head(dat))
```

As you can see, this dataset includes 3 columns:

- g1: a category variable with 3 levels, "a", "b", and "c"

- g2: another variable with 2 levels, "A","B", which are evenly balanced within
  each value of g1

- x: the value column, which will be statistically tested


## Tests

We will run a statistical test for the value `x` between "A" and "B" (column `g2`)
within each category of `g1`. For this, we will use dplyer's `group_by()` function
to divide data, and then run tests on subsets and each subset returns a data.frame.
This data.frame can be multi-row or single-row, and see how dplyr handles the return
results.

First, let's use the `summarize()` function to collapse the results.

```{r}
# this function return a single-row data.frame
my_test<-function(v, g) {
  res<-t.test(v ~ g)
  df<-data.frame(mean1=res$estimate[1], mean2=res$estimate[2], P=res$p.value)
  return(df)
}

dat %>%
  group_by(g1) %>%
  summarize(with(.data, my_test(x, g2) )) %>%
  kable()
```

As you can see, the functions do well, and return a new data.frame with
group name and the result from each group.

Let's also try a function returning multiple rows.


```{r}
# this function returns multiple-row data.frame
my_quantiles<-function(v) {
  probs<-seq(0,1,0.25)
  qt<-quantile(x, probs = probs)
  data.frame(quant=qt, prob=probs)
}
dat %>%
  group_by(g1) %>%
  summarize(with(.data, my_quantiles(x) )) %>%
  kable()
```

The results remain good but with deprecating warning, which asks me to replace
`summarize()` with `reframe()`, so let's try this new function.

```{r}
dat %>%
  group_by(g1) %>%
  reframe(with(.data, my_quantiles(x) )) %>%
  kable()
```

Great, everything looks good.

## Conclusions

We can combine the function `group_by()`, `summarize()` and a function returning a data.frame
to easily analyze data by groups via `dplyr`. When the returned data.frame is
multi-row, the function `summarize()` should be replaced with `reframe()`.

## References

- reframe(): https://dplyr.tidyverse.org/reference/reframe.html

- mutate() with multi-row results: https://stackoverflow.com/questions/73398676/dplyrmutate-when-custom-function-return-a-vector

