---
title: '[R] Use new_scale_xxx() function to add the same scale type in different ggplot
  layers'
author: Zhenguo Zhang
date: '2025-05-10'
slug: r-use-new-scale-xxx-function-to-add-the-same-scale-type-in-different-ggplot-layers
categories:
  - R
tags:
  - Bioinformatics
  - R
  - ggplot2
description: ''
featured_image: ''
images: []
comment: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, dpi=200, fig.width = 5, fig.height = 5, fig.align = "center")
```

In one ggplot figure, normally you can only use one scale for each aesthetic mapping. For example, if you use `scale_color_manual()` to set the color scale for a layer, you cannot use another `scale_color_manual()` for another layer, or
set the color scale more then once in the function `aes()`. However, you can use the `new_scale_color()` function from the `ggnewscale` package to add a new scale for the same aesthetic mapping in different layers.

In this post, I will showcase how to use the `new_scale_color()` function to add two different color scales in a ggplot figure. The first scale will be for a discrete variable (e.g., number of cylinders), and the second scale will be for a continuous variable (e.g., density level).

Load packages first.

```{r}
library(ggplot2)
library(ggnewscale)
```

Use the mtcars dataset for the example

```{r}
data(mtcars)
```


Create a plot with two color scales:
1. Points colored by 'cyl' (discrete)
2. Density contours colored by density level (continuous)


First, let's make a scatter plot of `mpg` vs `wt` with points colored by the number of cylinders (`cyl`). We will use the `geom_point()` function for this layer.

```{r}
plt <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  # First layer: Scatter plot colored by cylinders (discrete variable)
  geom_point(aes(color = factor(cyl)), size = 3) +  
  scale_color_discrete(name = "Cylinders")

plt
```

Set new scale for the next layer

```{r set-new-scale}
# Reset the color scale for the next layer
plt <- plt + new_scale_color()
```

Add a second layer: Density contours colored by density level (continuous variable)

```{r}
plt <- plt +
  geom_density_2d(aes(color = after_stat(level))) +  
  scale_color_viridis_c(name = "Density Level", option = "magma") +
  
  # Add labels and theme
  labs(title = "Dual Color Scales with new_scale_color()",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon") +
  theme_minimal()

plt
```


Here I demonstrated how to use the `new_scale_color()` function from the `ggnewscale` package, one can also use `new_scale_fill()` for fill aesthetics.
For other aesthetics, such as `size`, `shape`, etc., you can call
`new_scale("size")`, `new_scale("shape")`, etc. to add new scales.

To learn more, check the webpage https://eliocamp.github.io/ggnewscale/

Happy programming :smiley:
