---
title: '[R] Create ggplot histograms with y value being fractions'
author: "Zhenguo Zhang"
date: "2023-05-26"
slug: "r-create-ggplot-histograms-with-y-value-being-fractions"
categories:
- R
- Computing
tags:
- R
- Programming
- ggplot2
description: ''
featured_image: ''
images: []
comment: yes
---

```{r}
library(knitr)
opts_chunk$set(echo=T, warning=F, fig.width=7, fig.height=7)
library(ggplot2)
```

[ggplot2](https://ggplot2.tidyverse.org) has been a great tool for making
plots. Today, I would like to talk about histogram. It is straight forward
to make a histogram with `geom_histogram()`, but not so when one wants to use
fractions instead of counts as values on the y-axis, especially when data are
separated into groups. Let's dive in.

## Histogram without groups

First, let's generate some fake data.

```{r}
x<-1:10
y<-rep(c("a","b"), times=5)
dat<-data.frame(x,y)
```

This dataset contains only 10 values without duplicates, and they are divided
into two groups `a` and `b` labelled by variable `x`.

```{r}
plt<-ggplot(dat, aes(x=x)) + geom_histogram(binwidth = 1, color="red", fill="yellow")
plt + ggtitle("binwidth=1, count")
```

Here we specified binwidth=1, so there are 10 bins and 
each bin is expected to have only 1 count, and this
is exact what we observed. Now let's try a different binwidth, say 0.5. This will
create 20 bins, but half of them will be empty.

```{r}
plt<-ggplot(dat, aes(x=x)) + geom_histogram(binwidth = 0.5, color="red", fill="yellow")
plt + ggtitle("binwidth=0.5, count")
```

If we use fraction, we expect each fraction is being 0.1, because there are
only 10 values and each non-empty bin has 1 value. To do so, we can use
the function `after_stat(density*width)`, where the density is the density
value in each bin, and width is the binwidth.

```{r}
plt<-ggplot(dat, aes(x=x)) + geom_histogram(aes(y=after_stat(density*width)), binwidth = 0.5, color="red", fill="yellow")
plt + ggtitle("binwidth=0.5, fraction")
```

The plot shows up as expected and each fraction is 0.1.

Good, now let's introduce groups.

## Histogram with groups

```{r}
plt<-ggplot(dat, aes(x=x, fill=y)) + geom_histogram(aes(y=after_stat(density*width)), binwidth = 0.5, color="red")
plt + ggtitle("binwidth=0.5, fraction")
```

The two groups are filled in different colors, and the fraction in each bin is
0.2. This makes sense, because in each group, there are only 5 values and each
non-empty  bin has 1 value, accounting 0.2 of the total 5.


On the internet, there is another way to calculate fractions using `after_stat(count)`,
which works for cases without groups, but not for histograms with groups.
Here we can see it:

```{r}
plt<-ggplot(dat, aes(x=x, fill=y)) + geom_histogram(aes(y=after_stat(count/sum(count))), binwidth = 0.5, color="red")
plt + ggtitle("binwidth=0.5, calculated fraction")
```

As you can see, each bin has fraction 0.1. This is because the `sum(count)` gives
the total number, not the total per group.


## Density plot with groups

Another way to compare the distributions of multiple groups is to use density
plot, as shown below:

```{r}
plt<-ggplot(dat, aes(x=x, fill=y)) + geom_density(alpha=0.5)
plt + ggtitle("density plot")
```

Happy plotting :smile:

