---
title: '[R] tidy evaluation in ggplot2'
author: Zhenguo Zhang
date: '2023-10-23'
slug: r-tidy-evaluation-in-ggplot2
categories:
  - R
  - Visualization
tags:
  - ggplot2
  - R
description: ''
featured_image: ''
images: []
comment: yes
---

Since ggplot2 version 3.0.0, it started to support tidy evaluation, and use
the technique to replace `aes_` and `aes_string()`. In this post, I would like
to show how the syntax of ggplot2 changes accordingly.

```{r setup}
knitr::opts_chunk$set(echo=T, fig.align = "center", fig.width = 6, fig.height = 5, dpi=150, warning=FALSE)
```

```{r}
library(knitr)
library(ggplot2)
```

Let's start with an example of making a scatter plot with the
dataset `mtcars`.

```{r}
library(ggplot2)
plt<-ggplot(mtcars, aes(x=wt, y=mpg, color=factor(cyl))) + theme_bw()
plt<-plt+geom_point()
plt
```


But what if we create a function to make the plot and accept different variables for
the color aesthetics? 

```{r}
make_scatter_plot<-function(colorVar) {
  plt<-ggplot(mtcars, aes(x=wt, y=mpg, color=factor(colorVar))) + theme_bw()
  plt<-plt+geom_point()
  plt
}
make_scatter_plot("cyl")
```


As you can see, this doesn't work as expected: it actually created a new variable
which has only one value "cyl" for the `fill` aesthetics. What we need is to use the
`cyl` column of the `mtcars` dataset to fill the colors.

Here is tidy evaluation kicks in: we use `enquo()` to quote an input variable, and
then unquote it with `!!`, as shown in the function below:


```{r}
make_scatter_plot<-function(colorVar) {
  colorVar<-enquo(colorVar)
  plt<-ggplot(mtcars, aes(x=wt, y=mpg, color=factor(!!colorVar))) + theme_bw()
  plt<-plt+geom_point()
  plt
}
make_scatter_plot(cyl)
```


Here we used the function `ggplot2::enquo()` which delayed the evaluation of an expression,
and it will only be evaluated when prefixed with `!!`. This is the key feature of tidy evaluation.
With this function, we can call a different column to color it, for example:

```{r}
make_scatter_plot(gear)
```


Note that you can't put the variable name in a quote, such as "gear", and it would not work.

Another useful function is `ggplot2::vars()`, it can quasiquote more than one variables and 
be used in functions such as `facet_wrap()` and `facet_grid()`. Let's see an example:

```{r}
make_scatter_plot(gear) + facet_wrap(vars(am), labeller=label_both)

```


We can also put both color and facet variable in another function `make_scatter_plot_facet`
which calls `make_scatter_plot`. As you can see, the color variable need be `enquo` and unquote
(use `!!`) in this wrapping function too.

```{r}
make_scatter_plot_facet<-function(colorVar, facetVar) {
  #facetName<-quo_name(facetVar)
  facetVar<-enquo(facetVar)
  colorVar<-enquo(colorVar)
  make_scatter_plot(!!colorVar) + facet_wrap(vars(!!facetVar), labeller=label_both)
}

make_scatter_plot_facet(gear, am)

```


This is just the first post on tidy evaluation. More is coming.

Happy programming :smile:

## References

- tidy evaluation in ggplot2: https://www.tidyverse.org/blog/2018/07/ggplot2-tidy-evaluation/

- quasiquotation in R: https://adv-r.hadley.nz/quasiquotation.html
