---
title: '[R] tidy evaluation in ggplot2 (part 2)'
author: "Zhenguo Zhang"
date: "2023-11-11"
slug: "r-tidy-evaluation-in-ggplot2-part-2"
categories:
- R
- Software
- Visualization
tags:
- R
- ggplot2
- Programming
description: ''
featured_image: ''
images: []
comment: yes
---

```{r setup}
knitr::opts_chunk$set(echo=T, fig.align = "center", fig.width = 6, fig.height = 5, dpi=150, warning=FALSE)
```

```{r}
library(knitr)
library(ggplot2)
```

## How to use variables to refer columns in ggplot2

Let's start an example of making a scatter plot by using the
dataset `mtcars`.

```{r}
head(mtcars)
```

We can make a plot using the following commands

```{r}
plt<-ggplot(mtcars, aes(x=wt, y=mpg, color=factor(cyl))) + theme_bw()
plt<-plt+geom_point()
plt
```


If we create a function, how can we make the sample plot:

### use '{{}}'

```{r}
make_scatter_plot1<-function(colorVar) {
  plt<-ggplot(mtcars, aes(x=wt, y=mpg, color=factor({{colorVar}}))) + theme_bw()
  plt<-plt+geom_point()
  plt
}
make_scatter_plot1(cyl)
```

The embrace operator '⁠{{'⁠ is used to create functions that call other data-masking functions. It transports a data-masked argument (an argument that can refer to columns of a data frame) from one function to another.

### use .data

```{r}
make_scatter_plot2<-function(colorVar) {
  plt<-ggplot(mtcars, aes(x=wt, y=mpg, color=factor(.data[[colorVar]]))) + theme_bw()
  plt<-plt+geom_point()
  plt
}
make_scatter_plot2("cyl")
```

### use sym()

```{r}
make_scatter_plot3<-function(colorVar) {
  plt<-ggplot(mtcars, aes(x=wt, y=mpg, color=factor(!!sym(colorVar)) )) + theme_bw()
  plt<-plt+geom_point()
  plt
}
make_scatter_plot3("cyl")
```

### use quo()


```{r}
make_scatter_plot4<-function(colorVar) {
  plt<-ggplot(mtcars, aes(x=wt, y=mpg, color=factor(!!colorVar) )) + theme_bw()
  plt<-plt+geom_point()
  plt
}
myvar<-quo(cyl)
make_scatter_plot4(myvar)
```


## Key concepts

### defusing

When a piece of R code is defused, R doesn't return its value, but returns the
expression in a tree-like object to describe how to compute a value, which can
be evaluated later in a specific environment.

For example, one can use `expr()` to create a defused object.

```{r}
# return result
1+1
# return defused object
expr(1+1)
```

One application of using defused object is to resume its evaluation in data
mask, see the example below:

```{r}
e <- expr(mean(cyl))
eval(e, mtcars)
```

Some functions defuse their arguments, and when these function are called from
another function, its argument must be provided with  embrace operator \{\{, which
is a combination of `enquo()` and `!!`.

Defused expressions can also be created by using other functions
such as `sym()` and `call()`, for example:

```r
# create a symbol
vName<-"mtcars"
sym(vName)

# make a call
call("head", sym(vName))
```

There are two ways to defuse expressions, one can use
`expr()` to defuse one's own expression and use `enquo()` to defuse user-provided
expressions to your function.

Defusing is known as quoting in other frameworks.

-    The equivalent of `expr()` is [base::bquote()].

-    The equivalent of `enquo()` is [base::substitute()]. The latter returns a naked expression instead of a quosure.

-   There is no equivalent for enquos(...) but you can defuse dots as a list of naked expressions with eval(substitute(alist(...))).



### data-masking

Data masking means that a data.frame is inserted at the bottom of the current
chain of environment and it takes higher precedence than the user environment.

For example, if we assign a value to a variable, but this variable is invisible
in data-masking functions as below:

```{r}
mpg<-100
# the following function can't see it
dplyr::summarise(mtcars, mean(mpg))
```
As you can see, the function used the column `mpg` from the dataframe `mtcars`.

One way to disambiguate the user and mask environments is to use
[tidy eval pronouns](https://search.r-project.org/CRAN/refmans/rlang/help/.data.html).

```{r}
dplyr::summarise(mtcars, mean(.env$mpg))
```


## Summary

Basically, when one wants to evaluate a variable in a different context, he should
defuse the variable (using function such as `quo()` and `enquo()`) and 
then inject it (using `{{}}`, `!!`, `!!sym()`, etc) when evaluation is needed.

## References

- Using ggplot2 in packages: https://ggplot2.tidyverse.org/articles/ggplot2-in-packages.html

- Functional programming: https://dcl-prog.stanford.edu/tidy-eval-detailed.html

- Defuse and inject patterns: https://search.r-project.org/CRAN/refmans/rlang/html/topic-metaprogramming.html

- Data-masking programming: https://search.r-project.org/CRAN/refmans/rlang/html/topic-data-mask-programming.html
