---
title: '[R] Arrange images in a table'
author: "Zhenguo Zhang"
date: '2021-02-01'
slug: r-arrange-images-into-a-table
categories:
- R
- Computing
tags:
- Programming
- R Markdown
- R
---

```{r setup}
library(knitr)
library(kableExtra)
```

Sometimes, we need arrange images into a grid. I know two 
approaches for this. One is to manually fill table cells
with strings like ```![](path/to/img)```, and the other is
to use function *column_spec(column, image=vector_of_files)*
from [kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html).

In this post, I am going to show both approaches.

First, let's generate some figures.

## Generating figures

We will use the builtin dataset *mtcars*.

```{r generate-figs}
figFiles<-file.path(
                    c("scatter.png","hist.png",
                      "boxplot.png","density.png",
                      "barplot.png","line.png")
                    )
i=1;
width=5;
height=5;
unit="in";
dpi=300;
png(figFiles[i], width=width, height = height, units = unit, res = dpi)
with(mtcars, plot(mpg ~ cyl, main="scatter plot"))
tmp<-dev.off()
i<-i+1;
png(figFiles[i], width=width, height = height, units = unit, res = dpi)
with(mtcars, hist(cyl, main="hist plot"))
tmp<-dev.off()
i<-i+1;
png(figFiles[i], width=width, height = height, units = unit, res = dpi)
with(mtcars, boxplot(hp ~ as.factor(cyl), main="boxplot"))
tmp<-dev.off()
i<-i+1;
png(figFiles[i], width=width, height = height, units = unit, res = dpi)
with(mtcars, plot(density(wt), main="density plot"))
tmp<-dev.off()
i<-i+1;
png(figFiles[i], width=width, height = height, units = unit, res = dpi)
with(mtcars, barplot(table(gear), col="blue", main="barplot"))
tmp<-dev.off()
i<-i+1;
png(figFiles[i], width=width, height = height, units = unit, res = dpi)
with(mtcars, boxplot(hp ~ as.factor(cyl), main="boxplot"))
tmp<-dev.off()
```

## Put images into a table using manual fill

```{r img-mat-md}
md_img<-function(f)
{
  mdStr<-paste0(sprintf('![](%s)', f),'{width=100%}')
  #mdStr<-include_graphics(f)
  return(mdStr)
}
mat<-matrix(sapply(figFiles, md_img),nc=2, byrow = T)
kbl(mat, table.attr = 'style="width:80%;"', caption="Image markdown") %>%
  kable_styling(full_width = T, position="center") %>%
  column_spec(1:2, width="50%")
```

Here we created a help function 'md_img()' to embed each figure
file into a markdown string, and then we arrange the resulted
strings into a matrix with 3 columns. Finally, we use *kableExtra*
to render the table.

Note that to control the size of image, I used an attribute
"width=100%" when constructing the markdown string, which ensures
the image fit the cell size.

## Use column_spec to arrange images

```{r img-mat-cs}
mat<-data.frame(col1=rep("",3), col2="")
mat %>%
  kbl(caption="Image table by column_spec", booktabs = T, table.attr = 'style="width:80%;"') %>%
  kable_styling() %>%
  column_spec(1, image=spec_image(figFiles[1:3], width = 900, height=900)) %>%
  column_spec(2, image=spec_image(figFiles[4:6], width = 900, height = 900))
```

Here we first create a matrix filled with empty strings,
then using the *kable_styling* and *column_spec* to convert
the matrix to table and to add content to columns. I
also used *spec_image()* function to provide the images
to the columns instead of file names directly, because
the function can help to change image sizes here.

All done. Have fun!