---
title: '[R] format paragraph outout'
author: Zhenguo Zhang
date: '2019-09-26'
slug: r-format-paragraph-outout
categories:
  - Tips
  - R
tags:
  - R
  - Programming
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T, comment="", fig.align = "center")
```

When programming in R, one may print some messages such as usage
information. When it comes, the base function *[strwrap()](https://stat.ethz.ch/R-manual/R-devel/library/base/html/strwrap.html)*
becomes very convenient. It can wrap strings into equal-length
lines and add indentation/prefixes to lines. Here I will demonstrate
some features.

First, let's create a sample string by reading the first 5 lines
of the 'THANKS' document in R documentation.

```{r read-data}
# read 5 lines from the 'THANKS' document
x <- paste(readLines(file.path(R.home("doc"), "THANKS"),n=5), collapse = "\n")
print(x)
```

Let's try formatting the string to lines of 30-character long,
as shown below:

```{r simple}
writeLines(strwrap(x, width=30))
```

Now, let's indent the 1st line of each paragraph
by using the paramter *indent*.

```{r indent-5}
writeLines(strwrap(x, 30, indent = 5))
```

We can indent each line except the first line with the parameter
*exdent*:

```{r exdent-5}
writeLines(strwrap(x, 30, exdent = 5))
```
Now let's add a prefix to the start of the first paragraph using
the parameter *initial*:

```{r initial}
writeLines(strwrap(x, 30, initial = "THANKS: "))
```
We can add prefixes to the other lines using *prefix*:

```{r prefix}
writeLines(strwrap(x, 30, initial = "THANKS: ", prefix = " > "))
```

All done. Very useful function.

Note: 

1. any whitespaces in input strings are converted to spaces before
formatting, but double whitespaces are treated differently (see the
*strwrap* documentation for details).

2. when more than one string is given as inputs, in
default the result is to concatenate the strings into one text.
If one wants to keep each string's result separate, use the parameter
*simplify=F*.

Have R fun!
