---
title: '[Learning notes] R S4 classes'
author: Zhenguo Zhang
date: '2020-10-03'
slug: learning-notes-r-s4-classes
categories:
  - R
  - Computing
tags:
  - R
  - Programming
---

There are three kinds of classes in R, S3, S4, and RC (or R6). The S3 class is simply
created by adding a 'class' attribute to an object. The S4 class is stricter and need
formally created, but it is still different from the classes in other object-oriented
programming languages such as Java, in terms of class inheritance and method dispatch.
The RC/R6 classes are more like the classes in Java. One can find more information on
the differences at https://adv-r.hadley.nz/oo-tradeoffs.html.

Today I will share what I learned by showing an example of using S4 classes.

## Create S4 classes

All the methods used for creating S4 classes and methods are in the
R package [methods](https://rdrr.io/r/methods/methods-package.html).

To create a class, we use the *setClass* method.

```{r create}
# create a class called 'Shape', with two slots
Shape<-setClass("Shape", 
	slots=c(name="character",sides="numeric"),
	prototype=list(name=NA_character_, sides=NA_real_))
(shape1<-Shape(name="triangle", sides=3))
```

As you see above, we created a class called 'Shape' and assigned the
generator to a function with the same name 'Shape'. Then we create
a shape object 'shape1' with the generator.

## Add methods

Next we add some methods to the class. The code is as below:

```{r methods}
# set generic first
setGeneric(
  "plotIt",
  function(obj) {
    standardGeneric("plotIt")
  }
)
# then we add a method for this generic
setMethod("plotIt",
          "Shape",
          function(obj) {
            cat(sprintf("This is a %s with %d sides", obj@name, obj@sides))
          })
# and call this method on 'shape1' created above
plotIt(shape1)
```

As you see, we first created a generic function *plotIt* using
*setGeneric()*. Any class can implement this generic function by following
the same argument list. If the generic one wants to implement
has already existed or imported from other packages. This step
should be omitted.

Following it, we implemented a method for the class *Shape*
using *setMethod()*. Here this method just simply prints out
the information of the object.

Next we will add some subclasses and implement *plotIt* method
on them.

## Subclasses

To create subclasses, it is very simple: we still use *setClass()*
function, but we will use an argument *contains=parent*, which will
include all slots from parent class. Here we will create two subclasses,
triangle and square.

```{r subclasses}
triangle<-setClass("triangle",
                   contains="Shape",
                   slots=c(edgeLen="numeric"),
                   prototype = c(name="triangle",
                                 sides=3
                                # edgeLen=NA_real_
                                 )
                   )
square<-setClass("square",
                 contains = "Shape",
                 slots=c(width="numeric"),
                 prototype = c(name="square",
                               sides=4,
                               width=NA_real_))
# add plotIt method
setMethod("plotIt",
          signature("triangle"),
          function(obj) {
            cat(sprintf("This is a %s with %d sides [%s]", 
                obj@name, obj@sides, paste0(obj@edgeLen,collapse=",")));
            plot(1,pch=2, cex=10, axes=F, main="Triangle",xlab=NA, ylab=NA)
          }
          )

setMethod("plotIt",
          signature("square"),
          function(obj) {
            cat(sprintf("This is a %s with %d equal sides = %s",
                  obj@name, obj@sides, obj@width));
            plot(1,pch=0, cex=10, axes=F, main="Square",xlab=NA,ylab=NA)
          }
          )

# create two objects
triangle1<-triangle(edgeLen=c(1,1,1.5))
square1<-square(width=10)
# call the method
plotIt(triangle1);
plotIt(square1)
```

As you see above, we created two subclasses *triangle* and *square*,
and also add *plotIt* method for each (note that the plot doesn't use
the object's slot information, solely for demonstration purpose).
After that, we create two objects and called the method.

## S4 method dispatch

Here we have very simple class hierarchy, so it is easy to know
which method is called for each object. When class inheritance
levels go up, it becomes more complicated and may cause ambiguity
on which method to be called.

Briefly, when a method is called, it first tries to match the 
class of the actual arguments
in a call to the available methods collected for a generic, 
here ‘plotIt’.  If there is
a method defined for the exact same classes as in this call, that
method is used.  

If there is no method for the given arguments, then all possible
signatures are considered, including
the actual classes or to superclasses of the actual classes 
(including ‘"ANY"’).  The method having the least
distance from the actual classes is chosen; if more than one
method has minimal distance, one is chosen (the lexicographically
first in terms of superclasses) but a warning is issued.  All
inherited methods chosen are stored in another table, so that the
inheritance calculations only need to be done once per session per
sequence of actual classes.

Well, I'll stop here for this post. It shows how to create S4
classes and methods. One interesting in this topic may look into
a document by Bioconductor community at https://bioconductor.org/help/course-materials/2017/Zurich/S4-classes-and-methods.html.

