---
title: '[R] How to overlay points over boxplots'
author: Zhenguo Zhang
date: '2023-08-26'
slug: r-how-to-overlay-points-over-boxplots
categories:
  - R
  - Computing
  - Visualization
tags:
  - R
  - ggplot2
description: ''
featured_image: ''
images: []
comment: yes
---

```{r setup, include=F}
library(knitr)
opts_chunk$set(echo=T, warning=F, fig.width=7, fig.height=7, dpi=200, fig.align="center")
```

```{r}
library(ggplot2)
```

ggplot2 is a powerful tool to visualize data. Today I would like to show how
to make a boxplot and then overlay points.


## A pure boxplot

First, let's make a boxplot. We will use the data set `ToothGrowth` coming with
R.


```{r bx-plot}
dat<-ToothGrowth
dat$dose<-as.factor(dat$dose) # convert the dose to a factor
plt<-ggplot(dat, aes(x=dose, y=len, fill=supp)) + theme_bw()
plt<-plt + geom_boxplot()
plt
```


## add points with geom_jitter()

Now let's add points over the boxplots with the function
`geom_jitter()`. Here we use the function `position_jitter()`
to control the points' spreads.

```{r}
plt + geom_jitter(position = position_jitter(width=0.4), pch=24, size=2, alpha=0.6)
```

This doesn't look right, because the points are superimposed over both boxes in
each dose level; they are supposed to be separated by the factor 'supp'.

Now let's try a different function `position_jitterdodge()`


```{r}
plt + geom_jitter(position = position_jitterdodge(jitter.width=0.4), pch=24, size=2, alpha=0.6)
```


This looks good, right. Note that we used 'fill' to control the colors of both
boxplots and points, so the selected points by 'pch' should have a property of
fill.


## Add points with geom_point()


We can also use `geom_point()` plus `position_jitterdodge()` to make similar
plots, as below:

```{r}
plt + geom_point(position = position_jitterdodge(jitter.width=0.4), pch=24, size=2, alpha=0.6)
```

The function [position_jitterdodge()](ggplot2::position_jitterdodge()) provides jitter effects, and the degree of
spread can be tuned with the parameter `jitter.width`, default is 0.4. Let's try
a smaller number 0.2.


```{r}
plt + geom_point(position = position_jitterdodge(jitter.width=0.2), pch=24, size=2, alpha=0.6)
```

As you can see, the points are clustered more narrowly in the horizontal
direction.

## The categorical variable should be a factor

There is a caveat here. The variable used for specifying filling color
must be a factor, otherwise the points will not be grouped correctly.

Let's create a new variable `supp2`, which is a character version of `supp`,
and use `supp2` for color fill.


```{r}
dat$supp2<-as.character(dat$supp)
plt<-ggplot(dat, aes(x=dose, y=len, fill=supp2)) + theme_bw()
plt<-plt + geom_boxplot()
plt + geom_point(position = position_jitterdodge(jitter.width=0.4), pch=24, size=2, alpha=0.6)
```

As you can see, the points are spreaded out, even though the same parameter
`jitter.width` was given. Let's try a smaller value 0.2


```{r}
plt + geom_point(position = position_jitterdodge(jitter.width=0.2), pch=24, size=2, alpha=0.6)
```

This helps a bit.


## Conclusions

To make boxplot with superimposed points, one need use the function
`position_jitterdodge()` to specify point positions, with this both
`geom_point()` and `geom_jitter()` can be used.

Also the group variables used for specifying colors or filling colors
should be a factor, not character; otherwise some unexpected behavior may
happen.

Happy Programming :smile: :+1:

## References

- ggplot2 boxplot quick start: http://www.sthda.com/english/wiki/ggplot2-box-plot-quick-start-guide-r-software-and-data-visualization
