---
title: '[R] How to use ggpattern to add patterns to ggplot2 plots'
author: Zhenguo Zhang
date: '2026-05-22'
slug: r-how-to-use-ggpattern
categories:
  - R
tags:
  - R
  - ggplot2
  - visualization
  - ggpattern
description: 'A guide on how to use the ggpattern package to add various patterns to ggplot2 plots, including installation, basic usage, and customizing pattern parameters.'
comment: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
library(ggplot2)
library(ggpattern)
library(dplyr)
```

Adding patterns to plots is a great way to improve accessibility (making plots colorblind-friendly) and to add an extra dimension of information. The `ggpattern` package provides a rich set of tools to achieve this in `ggplot2`.

## Basic Example with mtcars

Let's start with a barplot using the `mtcars` dataset. We will map `cyl` to the x-axis, `mpg` to the y-axis, use `gear` for the fill color, and `am` for the pattern.

```{r basic-example}
# Prepare data
df <- mtcars %>%
  mutate(
    cyl = factor(cyl),
    gear = factor(gear),
    am = factor(am, labels = c("Automatic", "Manual"))
  )

# Create barplot
ggplot(df, aes(x = cyl, y = mpg, fill = gear)) +
  geom_bar_pattern(
    aes(pattern = am),
    stat = "identity",
    position = "dodge",
    color = "black",
    pattern_fill = "black",
    pattern_spacing = 0.02,
    pattern_key_scale_factor = 1.5
  ) +
  scale_pattern_manual(values = c("stripe", "crosshatch")) +
  labs(
    title = "MPG by Cylinders, Gear, and Transmission",
    subtitle = "Patterns representing transmission type",
    x = "Cylinders",
    y = "Miles Per Gallon (MPG)",
    fill = "Gears",
    pattern = "Transmission"
  ) +
  theme_minimal()
```

## Exploring Key Parameters

The `ggpattern` package allows for fine-grained control over how patterns are rendered. Here are some of the most important parameters:

### 1. pattern_key_scale_factor

This parameter scales the pattern density specifically in the legend. Without it, patterns in the legend can often be hard to see.
Reduce the value will make it easier fit into the legend keys.

```{r scale-factor}
p_base <- ggplot(df, aes(x = cyl, y = mpg, fill = gear)) +
  geom_bar_pattern(
    aes(pattern = am),
    stat = "identity",
    position = "dodge",
    color = "black",
    pattern_spacing = 0.02
  ) +
  theme_minimal() +
  labs(title = "Default vs Scaled Legend")

# Default legend
p1 <- p_base + labs(subtitle = "Default (key_scale_factor = 1)")

# Scaled legend
p2 <- p_base + 
  geom_bar_pattern(
    aes(pattern = am),
    stat = "identity",
    position = "dodge",
    color = "black",
    pattern_spacing = 0.02,
    pattern_key_scale_factor = 0.5
  ) +
  labs(subtitle = "Scaled (key_scale_factor = 0.5)")

p1
p2
```

### 2. pattern_spacing and pattern_density

`pattern_spacing` controls the distance between pattern elements (lower = denser), while `pattern_density` controls the coverage (higher = denser).

```{r spacing-density}
# Spacing effect
ggplot(df, aes(x = cyl, y = mpg, fill = gear)) +
  geom_bar_pattern(
    aes(pattern = am),
    stat = "identity",
    position = "dodge",
    pattern_spacing = 0.2, # Wider spacing
    pattern_key_scale_factor = 1.5
  ) +
  labs(title = "Effect of pattern_spacing = 0.5 (Sparser)") +
  theme_minimal()

# Density effect
ggplot(df, aes(x = cyl, y = mpg, fill = gear)) +
  geom_bar_pattern(
    aes(pattern = am),
    stat = "identity",
    position = "dodge",
    pattern_density = 0.8, # Higher density
    pattern_key_scale_factor = 1.5
  ) +
  labs(title = "Effect of pattern_density = 0.8 (Denser)") +
  theme_minimal()
```

### 3. pattern_fill

`pattern_fill` sets the color of the pattern elements themselves.

```{r pattern-fill}
ggplot(df, aes(x = cyl, y = mpg, fill = gear)) +
  geom_bar_pattern(
    aes(pattern = am),
    stat = "identity",
    position = "dodge",
    pattern_fill = "white", # White pattern lines
    pattern_density = 0.5, # make it bigger so that the color is easier to see
    pattern_spacing = 0.02,
    pattern_key_scale_factor = 1.5
  ) +
  labs(title = "Effect of pattern_fill = 'white'") +
  theme_minimal()
```

### 4. pattern_shape

For geometric patterns like `regular_polygon`, `pattern_shape` controls the number of sides.

```{r pattern-shape}
ggplot(df, aes(x = cyl, y = mpg, fill = gear)) +
  geom_bar_pattern(
    aes(pattern_shape=gear),
    pattern = "regular_polygon",
    stat = "identity",
    position = "dodge",
    pattern_size = 0.5, # stroke line width
    pattern_density = 0.5, # increase the shape size
    pattern_spacing = 0.03,
    pattern_key_scale_factor = 1.5
  ) +
  scale_pattern_shape_manual(values = c("3"="circle", "4"="convex6", "5"="square")) + # manually set shapes
  labs(title = "Effect of mapping pattern_shape to gear") +
  theme_minimal()
```

## Troubleshooting

If you encounter issues installing or loading `ggpattern` on Linux, you may need to install the following system dependencies:

```bash
sudo apt-get install libudunits2-dev
sudo apt install libproj22
sudo apt-get install libgdal-dev
```
