---
title: '[Tutorial] How to create and tune ridgeline density plot using ggridges'
author: Zhenguo Zhang
date: '2023-07-01'
slug: tutorial-how-to-create-and-tune-ridges-density-plot-using-ggridges
categories:
  - R
  - Software
  - Visualization
tags:
  - R Markdown
  - ggplot2
description: ''
featured_image: ''
images: []
comment: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T, warning=F)
```

Today, I would like to introduce how to use the package
[ggridges](https://cran.r-project.org/web/packages/ggridges/index.html)
to create a ridgeline density plot. Let's start.

## Install the package

```r
install.packages("ggridges")
```

## First try

```{r}
library(ggplot2)
library(ggridges)
library(dplyr)
plt<-ggplot(iris, aes(x=Sepal.Width, y=Species, fill=Species))
plt<-plt+theme_bw()
plt+geom_density_ridges()
```

Looks good, right.

## Add data points

To do so, we will add rug lines at the bottom of each distribution, by
turn on the option `jittered_points`.

```{r}
plt<-plt+geom_density_ridges( # add rug points
    jittered_points = TRUE,
    position = position_points_jitter(width = 0.05, height = 0.01, yoffset = -0.1),
    point_shape = '|', point_size = 2, point_alpha = 0.8, alpha = 0.7, point_color="blue")
plt
```

Parameter explanation:

- position_points_jitter() is used to determine where the rug lines are, here `width`
  and `height` gives the range for the lines to move randomly, and yoffset tells to
  move all lines downwards a certain amount.

- point_shape, point_size, point_alpha, point_color all determine the appearance of the
  rug lines.

## Add the number of data points for each distribution

Get counts

```{r}
pointCnt<-iris %>% group_by(Species) %>% count()
```

Update the plot with `geom_text()` to add counts.

```{r}
plt+geom_text(data=pointCnt, aes(x=4.5, y=Species, label=n))
```

Let's move the count number a bit to the top

```{r}
plt<-plt+geom_text(data=pointCnt, aes(x=4.5, y=Species, label=n),
  nudge_y = 0.95, vjust=1, size=7/.pt, color="red")
plt
```

Here we use the parameter `nudge_y` to shift the y value upwards
a certain amount and set text size and color.

It looks good, isn't it?

## References

One can find more examples of how to tune the ridgeline density plot
at the following pages:

- https://r-charts.com/distribution/ggridges/

- https://wilkelab.org/ggridges/articles/introduction.html

- https://cran.r-project.org/web/packages/ggridges/vignettes/gallery.html

Happy programming :smile: !

