---
title: '[R] How to create errorbars with overlaid points using ggplot'
author: Zhenguo Zhang
date: '2025-04-26'
slug: r-how-to-create-an-error-barplot-with-overlaid-points-using-ggplot
categories:
  - R
  - Software
tags:
  - R
  - ggplot2
  - plot
description: ''
featured_image: ''
images: []
comment: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 5, fig.height = 5, dpi=200)
```

```{r}
library(ggplot2)
library(dplyr)
```

Sometimes you may want to create a plot with the following features:

- a point to indicate the mean of a group
- error bars to indicate the standard deviation of the group
- and each group may have subgroups, which are represented by different colors.

In this post, I will show you how to create such a plot using the `ggplot2` package in R.

We will use the builtin `mtcars` dataset as an example. And we need to
compute the following variables for later use:

- The mean mpg for each group of `cyl` (number of cylinders) and `gear`` (number of gears), here `cyl` is the main group and `gear` is the subgroup.

```{r}
# Load the mtcars dataset
data(mtcars)
# Compute the mean and standard deviation of mpg for each group
mtcars_summary <- mtcars %>%
  group_by(cyl, gear) %>%
  summarise(mean_mpg = mean(mpg), sd_mpg = sd(mpg)) %>%
  ungroup()
# replace the NA values in sd_mpg with 1
mtcars_summary$sd_mpg[is.na(mtcars_summary$sd_mpg)] <- 1
# convert group variables into factors
mtcars_summary$cyl <- factor(mtcars_summary$cyl)
mtcars_summary$gear <- factor(mtcars_summary$gear)
```


## Create the plot - first try

Now we can create the plot using `ggplot2`. We will use the `geom_point()` function to create the points, and the `geom_errorbar()` function to create the error bars. We will also use the `aes()` function to specify the aesthetics of the plot.

```{r}
# Create the plot
plt <- ggplot(mtcars_summary, aes(x = cyl, y = mean_mpg, color = gear)) +
  geom_point(size = 3) + # add points
  geom_errorbar(aes(ymin = mean_mpg - sd_mpg, ymax = mean_mpg + sd_mpg), width = 0.2) + # add error bars
  labs(x = "Number of Cylinders", y = "Mean MPG", color = "Number of Gears") + # add labels
  theme_minimal() + # use a minimal theme
  theme(legend.position = "top") # move the legend to the top
plt
```


Well, it is working, but the problem is that the error bars and points are all
aligned at the same position of x-axis. This is not what we want. We want the
subgroups to be separated by a small distance.

## Create the plot - second try

To separate the subgroups, we can use the `position_dodge()` function. This function will move the points and error bars to the left and right, so that they are not overlapping.

```{r}
pd <- position_dodge(width = 0.5)
# Create the plot with position_dodge
plt <- ggplot(mtcars_summary, aes(x = cyl, y = mean_mpg, color = gear)) +
  geom_point(size = 3, position = pd) + # add points with position_dodge
  geom_errorbar(aes(ymin = mean_mpg - sd_mpg, ymax = mean_mpg + sd_mpg), width = 0.2, position = pd) + # add error bars with position_dodge
  labs(x = "Number of Cylinders", y = "Mean MPG", color = "Number of Gears") + # add labels
  theme_minimal() + # use a minimal theme
  theme(legend.position = "top") # move the legend to the top
plt
```

Cool. Isn't it?

The only difference is that we added the `position = pd` argument to the `geom_point()` and `geom_errorbar()` functions. This tells `ggplot2` to use the `position_dodge()` function to separate the subgroups.


## Conclusion

In this post, we learned how to create a plot with error bars and overlaid points using the `ggplot2` package in R. We also learned how to separate the subgroups using the `position_dodge()` function.

If you want to learn more about the function `position_dodge()`, you can check an
excellent post [here](https://stackoverflow.com/questions/34889766/what-is-the-width-argument-in-position-dodge/35102486#35102486).


Happy programming! :smiley:
